Application of the XT3D Multi-Point Flux Approximation to Vertically Staggered Grids in MODFLOW 6 to Improve Accuracy of Simulated Flows in Steeply Dipping Layers

### 2022-09-14 latex/article.tex
Added a thought about the conductivity contrast - a possible case to try
Inserted some figures with preliminary results into the Results and Discussion

XXXXX Random thoughts:
\begin{itemize}
	\item Vertical offset happens even on a completely rectilinear grid if the water table elevation varies in horizontally adjacent cells. Doesn't seem like the same kind of issue, but think on it.
	\item What happens with a head gradient perpendicular to the dip of the channel (across rather than along the channel)? Preliminary results suggest cross-connections and xt3d help.
	\item Difference between truly single-layer model of channel vs embedded single-layer channel -- no vertical connections vs vertical connections that are no-flux. Effect on specific discharge calculations?
\end{itemize}

@@

Figure \ref{fig:schem-conn-area-flux}b shows the four cells as they would be configured for a vertically offset DIS or DISV grid in a two-layer, cross-sectional MODFLOW 6 model of the channel. The cells have horizontal tops and bottoms and can therefore follow the dip of the channel only on average, in what \cite{bardot2022} refer to as ``staircase" fashion. The pattern of grid connectivity is analogous to that in figure \ref{fig:schem-conn-area-flux}a, but it is not based strictly on which cell face overlap. Rather, a cell is connected to each cell in the same layer with which it has an overlapping face and with each cell in an adjacent layer with which it shares a horizontal face. Significantly, cells that have overlapping vertical faces but are in different layers (e.g., cells 1A and 2B) are not connected. In the conductance-based formulation for flow, the flow between two cells in the same layer of a DIS or DISV grid (XXXXX or DISU that's not vertically staggered, but do we want to mention that here? XXXXX) is proportional to the difference in heads computed at the two cell centers and a conductance based on an effective hydraulic conductivity for the connection between the cells. If the two cells are fully saturated and have the same hydraulic conductivity and thickness, as in this example, the cell interface area used to calculate the conductance is effectively based on the full cell thickness, as shown in figure \ref{fig:schem-conn-area-flux}b. Also, the length on which the conductance is based is the horizontal distance, not the straight-line distance, between the cell centers.

@@

\noindent where $h$ is head in meters and $x$ and $z$ are the horizontal and vertical model coordinates, respectively, in meters. Flow across the channel (XXXXX seems like this would be good to include XXXXX) is simulated using the modified head function
\begin{equation}
\label{eqn:head_analyt_across}
h = x \sin \theta - z \cos \theta,
\end{equation}
which represents a unit head gradient perpendicular to the long axis of the channel. In both cases, the hydraulic conductivity of the channel is set to 1 m/d (meter per day) so that the analytical flow solution is a groundwater flux of 1 m/d either along or across the channel. The conductivity of the domain surrounding the channel varies according to the scenario being investigated. In the homogeneous limit, the conductivity of the domain is set equal to that of the channel. In the highly heterogeneous limit, the conductivity of the domain is set much lower than that of the channel to effectively isolate the channel hydraulically.

@@

XXXXX Describe conversion XXXXX

@@

XXXXX Idea: See how heterogeneity affects the error -- series of runs with varying permeability contrast between channel and surrounding domain (from 1:1 to very large), producing a plot along the lines of figs. 3d and 4d in \cite{bardot2022} XXXXX

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[scale=0.5]{../figures/disu-x-cc-head.png}
	\caption{Cross-connections are good. XT3D is good.}
	\label{fig:mymodel-head}
	\end{center}
\end{figure}

>>>>

XXXXX Random thoughts:
\begin{itemize}
	\item Vertical offset happens even on a completely rectilinear grid if the water table elevation varies in horizontally adjacent cells. Doesn't seem like the same kind of issue, but think on it.
	\item What happens with a head gradient perpendicular to the dip of the channel (across rather than along the channel)? Preliminary results suggest cross-connections and xt3d help.
	\item Could also try making the channel less permeable than the domain; semi-confining unit.
	\item Difference between truly single-layer model of channel vs embedded single-layer channel -- no vertical connections vs vertical connections that are no-flux. Effect on specific discharge calculations?
\end{itemize}

@@

Figure \ref{fig:schem-conn-area-flux}b shows the four cells as they would be configured for a vertically offset DIS or DISV grid in a two-layer, cross-sectional MODFLOW 6 model of the channel. The cells have horizontal tops and bottoms and can therefore follow the dip of the channel only on average, in what \cite{bardot2022} refer to as ``staircase" fashion. The pattern of grid connectivity is analogous to that in figure \ref{fig:schem-conn-area-flux}a, but it is not based strictly on which cell faces overlap. Rather, a cell is connected to each cell in the same layer with which it has an overlapping face and with each cell in an adjacent layer with which it shares a horizontal face. Significantly, cells that have overlapping vertical faces but are in different layers (e.g., cells 1A and 2B) are not connected. In the conductance-based formulation for flow, the flow between two cells in the same layer of a DIS or DISV grid (XXXXX or DISU that's not vertically staggered, but do we want to mention that here? XXXXX) is proportional to the difference in heads computed at the two cell centers and a conductance based on an effective hydraulic conductivity for the connection between the cells. If the two cells are fully saturated and have the same hydraulic conductivity and thickness, as in this example, the cell interface area used to calculate the conductance is effectively based on the full cell thickness, as shown in figure \ref{fig:schem-conn-area-flux}b. Also, the length on which the conductance is based is the horizontal distance, not the straight-line distance, between the cell centers.

@@

\noindent where $h$ is head in meters and $x$ and $z$ are the horizontal and vertical model coordinates, respectively, in meters. Flow across the channel (XXXXX seems like this would be good to include XXXXX) is simulated using the modified head function
\begin{equation}
\label{eqn:head_analyt_across}
h = x \sin \theta - z \cos \theta,
\end{equation}
which represents a unit head gradient perpendicular to the long axis of the channel. In both cases, the hydraulic conductivity of the channel is set to 1 m/d (meter per day) so that the analytical flow solution is a groundwater flux of 1 m/d either along or across the channel. The conductivity of the domain surrounding the channel varies according to the scenario being investigated. In the homogeneous limit, the conductivity of the domain is set equal to that of the channel. In the highly heterogeneous limit, the conductivity of the domain is set much lower than that of the channel to effectively isolate the channel hydraulically. XXXXX Could also try making the channel less permeable than the domain; semi-confining unit XXXXX

@@

Jupyter notebooks for both test problems are available in the Supporting Information that accompanies this paper. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting} to assist in setting up input for and processing results from the MODFLOW 6 simulations. The second test problem is formulated by starting with a DIS grid, converting it to an equivalent DISU grid, and adding cross-connections. A Python script for converting and modifying the grid is included.

@@

\begin{figure}[H]
\centering
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-s-nocc-head.png}
	\caption{standard conductance-based formulation}
	\label{fig:disu-s-nocc-head}
\end{subfigure}
\hfill
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-x-nocc-head.png}
	\caption{XT3D}
	\label{fig:disu-x-nocc-head}
\end{subfigure}
\caption{Specific discharge and head for a vertically offset grid. The news is all bad; XT3D doesn't really help. (disu\_approach model, 5x11 grid without cross-connections, $30^{\circ}$ incline, set up for flow along the channel)}
\label{fig:figures}
\end{figure}

XXXXX Idea: See how heterogeneity affects the error -- series of runs with varying permeability contrast between channel and surrounding domain (from 1:1 to very large), producing a plot along the lines of figs. 3d and 4d in \cite{bardot2022}. Will need to use the transect\_benchmark model for this, as we probably should for all of the final simulations. XXXXX

\begin{figure}[H]
\centering
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-s-nocc-cf-head.png}
	\caption{standard conductance-based formulation}
	\label{fig:disu-s-nocc-cf-head.}
\end{subfigure}
\hfill
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-x-nocc-cf-head.png}
	\caption{XT3D}
	\label{fig:disu-x-nocc-cf-head}
\end{subfigure}
\caption{Specific discharge and head for a vertically offset grid and boundary heads set up for crossflow. The standard formulation is a fail, but XT3D nails it, even without cross-connections. (disu\_approach model, 5x11 grid without cross-connections, $30^{\circ}$ incline, set up for flow across the channel)}
\label{fig:figures}
\end{figure}

@@

\begin{figure}[H]
\centering
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-s-cc-head.png}
	\caption{standard conductance-based formulation}
	\label{fig:disu-s-cc-head}
\end{subfigure}
\hfill
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-x-cc-head.png}
	\caption{XT3D}
	\label{fig:disu-x-cc-head}
\end{subfigure}
\caption{Specific discharge and head for a vertically staggered grid with cross-connections. The news is much better; the standard formulation isn't horrible, and XT3D nails it. (disu\_approach model, 5x11 grid with cross-connections, $30^{\circ}$ incline, set up for flow along the channel)}
\label{fig:figures}
\end{figure}

\begin{figure}[H]
\centering
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-s-cc-cf-head.png}
	\caption{standard conductance-based formulation}
	\label{fig:disu-s-cc-cf-head}
\end{subfigure}
\hfill
\begin{subfigure}{0.4\textwidth}
	\includegraphics[width=\textwidth]{../figures/disu-x-cc-cf-head.png}
	\caption{XT3D}
	\label{fig:disu-x-cc-cf-head}
\end{subfigure}
\caption{Specific discharge and head for a vertically staggered grid with cross-connections and boundary heads set up for crossflow. The standard formulation does pretty well, and XT3D nails it again. (disu\_approach model, 5x11 grid with cross-connections, $30^{\circ}$ incline, set up for flow across the channel)}
\label{fig:figures}
\end{figure}
